---
title: "Lake Water Quality Analysis"
author: "Orion and Matthew Ross"
date: "2/23/2022"
output: html_document
---



```{r setup, include=FALSE}

#install.packages(c("RApiSerialize", "LAGOSNE", 'USAboundaries'))

library(tidyverse) # Tidy packages
library(sf) #Spatial package that can read and create shapefiles 
library(mapview) #Interactive maps
library(LAGOSNE) #Lots and lots of clean lake data
library(USAboundaries) #USA states and counties
library(lubridate) #For dealing with date and time
library(systemfonts)
```


# LAGOS Analysis


## Loading in data


### First download and then specifically grab the locus (or site lat longs)
```{r data-read}

#Load in lagos
lagos <- lagosne_load(fpath = "/Users/orioncr/Downloads/LAGOSNE/LAGOSNE/data_1.087.3.qs")

#Grab the lake centroid info
lake_centers <- lagos$locus

# Make an sf object 
spatial_lakes <- st_as_sf(lake_centers,coords=c('nhd_long','nhd_lat'),
                          crs=4326)

#Grab the water quality data
nutr <- lagos$epi_nutr

#Look at column names
#names(nutr)
```

### Subset columns nutr to only keep key info that we want


```{r}
clarity_only <- nutr %>%
  select(lagoslakeid,sampledate,chla,doc,secchi) %>%
  mutate(sampledate = as.character(sampledate) %>% ymd(.))

```


### Keep sites with at least 200 observations 

```{r}

#Look at the number of rows of dataset
#nrow(clarity_only)

chla_secchi <- clarity_only %>%
  filter(!is.na(chla),
         !is.na(secchi))

# How many observatiosn did we lose?
# nrow(clarity_only) - nrow(chla_secchi)


# Keep only the lakes with at least 200 observations of secchi and chla
chla_secchi_200 <- chla_secchi %>%
  group_by(lagoslakeid) %>%
  mutate(count = n()) %>%
  filter(count > 200)


```


### Join water quality data to spatial data

```{r}
spatial_200 <- inner_join(spatial_lakes,chla_secchi_200 %>%
                            distinct(lagoslakeid,.keep_all=T),
                          by='lagoslakeid')


```

### Mean Chl_a map

```{r}
### Take the mean chl_a and secchi by lake

mean_values_200 <- chla_secchi_200 %>%
  # Take summary by lake id
  group_by(lagoslakeid) %>%
  # take mean chl_a per lake id
  summarize(mean_chl = mean(chla,na.rm=T),
            mean_secchi=mean(secchi,na.rm=T)) %>%
  #Get rid of NAs
  filter(!is.na(mean_chl),
         !is.na(mean_secchi)) %>%
  # Take the log base 10 of the mean_chl
  mutate(log10_mean_chl = log10(mean_chl))

#Join datasets
mean_spatial <- inner_join(spatial_lakes,mean_values_200,
                          by='lagoslakeid') 

#Make a map
mapview(mean_spatial,zcol='log10_mean_chl')
```


# Class work

## 1) What is the correlation between Secchi Disk Depth and Chlorophyll a for
sites with at least 200 observations?

- Here, I just want a plot of chla vs secchi for all sites 

```{r}
#Your code here
ggplot(chla_secchi_200,aes(x=chla,y=secchi)) + 
  geom_point() +
  ggthemes::theme_few() + 
  labs(title = "Chlorophyll a vs Secchi Disk Depth", x = "Chlorophyll a",
       y = "Secchi Disk Depth ") +
  ggthemes::scale_color_few()
```


## Why might this be the case? 

Secci disk depth appears to decrease as Chlorophyll a increases.
Increased concentrations of chlorophyll a indicate more algal growth likely
increasing the turbidity of the water.

## 2) What states have the most data? 

The top 5 in order of observation count are Minnesota, Michigan, Wisconsin, New York and Illinois

### 2a) First you will need to make a lagos spatial dataset that has the total 
number of counts per site.

```{r}
## Your code here
counts_per_site <- lake_centers %>%
  group_by(lagoslakeid) %>%
  mutate(count = n()) %>%
  select(lagoslakeid,nhd_long,nhd_lat, count, state_zoneid)

spatial_site_count <- inner_join(spatial_lakes,counts_per_site %>%
distinct(lagoslakeid,.keep_all=T),
by='lagoslakeid')
```


### 2b) Second, you will need to join this point dataset to the us_boundaries 
data. 

```{r}
## Your code here
states = us_states() %>%
  st_transform(2163)

spatial_counts_per_site <- st_as_sf(counts_per_site,coords=c('nhd_long','nhd_lat'),
                          crs=4326) %>%
  st_transform(2163)

join_states_counts <- st_join(spatial_counts_per_site,states)

```


### 2c) Then you will want to group by state and sum all the observations in that
state and arrange that data from most to least total observations per state. 

```{r}
## Your code here. 
state_sums <- join_states_counts %>%
  group_by(name) %>%
  mutate(observations = sum(count)) %>%
  distinct(name, .keep_all = TRUE) %>%
  arrange(desc(observations))

mapview(state_sums, zcol='observations')
 
```

##3 Is there a spatial pattern in Secchi disk depth for lakes with at least 200 
observations?

A strong spatial pattern isn't apparent. Upon closer analysis, I observed clustering of high secci disk depth values in the Maine region and clustering of low and mid values in the Michigan to Minnesota region.
```{r}
## Your code here
secchi <- clarity_only %>%
  filter(!is.na(secchi))

# Keep only the lakes with at least 200 observations of secchi
secchi_200 <- secchi %>%
  group_by(lagoslakeid) %>%
  mutate(count = n()) %>%
  filter(count > 200)

secchi_spatial_200 <- inner_join(spatial_lakes,secchi_200 %>%
                            distinct(lagoslakeid,.keep_all=T),
                          by='lagoslakeid') %>%
  arrange(desc(secchi))

mapview(secchi_spatial_200, zcol = 'secchi')
```


